---
title: "Data Storytelling"
author: "Ted Laderas"
date: " `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
library(here)
library(readr)
library(tidyverse)
cvd_patient <- readr::read_csv(here("data/cvd_patient.csv"))
```

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono")
)
```

# Overview

- Adjust your attitude
- Knowing Your Audience/Building a Story
- Some visuals for presenting
- Highlight your message visually

---
class: center, middle
# Adjust Your Attitude

---
# Motivation: Exploratory versus Explanatory

- Exploratory analysis: exploring and understanding the data, conducting the analysis
- Explanatory analysis: explaining your findings from your analysis in a coherent narrative that leads to a call to action

---
# You are not your audience

- Think about what matters to your stakeholders
- Adjust your language
- Adjust your metrics

---
# Use Plain Language (Allison Stiter)

| Instead of...                | Try...                                                                                         |
|---------------------------|---------------------------------------------------------------------------------------------|
| Statistically Significant | "If we did this experiment again, it  would be unlikely to come up with a different result" |
| Null Result               | "We haven't gotten enough data to answer your question. More data may help."                |
| Confidence Interval       | "We're Pretty sure the true answer is between this and that."                               |
| p-value                   | Maybe skip talking about p-values at all?                                                   |
---
class: center, middle
# Know Your Audience and Build Your Story

---
# Three Questions before you even make slides and figures

--
- For whom?
--

- What is your call to action?
--

- What are your findings? How will you communicate them using the available data?

.footnote[Cole Nussbaumer Naflic, [Storytelling with Data](http://www.storytellingwithdata.com/books)]

---
# Who is your audience?

- What are their values?
- What are their strengths?
- How should you tailor your message?

---
# What is your call to action?

- What action do you want your audience to take?
- Example: For a pilot project of LACE:
  - What were your results on your pilot project?
  - What would it take to bring the project to the larger organization?

---

# Storyboarding Activity (15 Minutes)

You are part of a team that has done a pilot implementation of LACE in order to assess the viability. You are presenting to your sponsors in your organization about rolling out LACE to the entire organization.

Activity: Thinking about the sponsor of your analytic effort:

1. What is your call to action? (What are you asking for?)
1. What were your findings on the pilot? 
1. How will you connect them?

---
# What data do you need to tell your story?

- What datasets/tables/metrics?
- What numbers are meaningful?
- Should you subset the data?
- How should you present it visually?

---
class: center, middle

# Visuals for Presenting

---
# Types of visuals

- Numbers
- Barplots
- Heatmaps

---
# Just use the number

Tables do require processing. It may be worth it to just use the number as the visual.

---
# Barplots


---
# Heatmaps

May be better than straight tables!

---
class: center, middle
# Highlighting your message

---
# Effective Visual Communication

Focus on three techniques

- Preattentive Attributes
- Decluttering your plot using `theme()`
- Annotating your graph

---
# Preattentive Attributes

Ways to highlight properties of graphs even before people are even "paying attention".

- color
- contrast

---
# Let's summarize the data

Summarizing by race and cardiovascular disease

Using `group_by()/summarize()`:

```{r}
cvd_race_count <- cvd_patient %>% 
  mutate(race=factor(race)) %>% 
  group_by(race, cvd) %>% summarize(count=n()) #<<
```
---
# CVD Race Count

```{r echo=FALSE}
cvd_race_count
```

---
# Context: Ways to highlight parts of a graph

`scale_fill_manual()` is your friend with categorical data. Map your categories to colors manually, only highlight the categories you want to emphasize.

First add a column called `highlight` that emphasizes the group of interest. 

```{r}
fullset <- cvd_patient %>% 
  filter(cvd == "Y") %>%
  mutate(race=factor(race)) %>% 
  #count number of patients in each race category
  group_by(race) %>% summarize(count=n()) %>%
  #reorder race variable by the count in the category
  mutate(race=fct_reorder(race, count)) %>%
  mutate(highlight = #<<
           ifelse(race == "Black/AfAm", "Y", "N")) #<<
```
---
## Highlight Column

There's a new column called `highlight`:

```{r echo=FALSE}
DT::datatable(fullset)
```

---
## Highlighting the data

```{r}
out_plot <- ggplot(data=fullset, aes(x=race, y=count, fill=highlight)) + 
  geom_col() + 
  geom_text(aes(label=race, y=count+300)) +
  
  # Use the highlight variable with scale_fill_manual
  scale_fill_manual(values = c("Y"="blue", "N"="grey")) + #<<
  
  # Remove legend and x-axis information
  theme(legend.position = "none", #<<
        axis.text.x = element_blank()) + #<< 
    ggtitle("African Americans are 3rd in Number of\n Cardiovascular Cases for our cohort") #<<

```

---
# Resulting Graph

```{r echo=FALSE}
out_plot
```

---
# Reduce Clutter

Reduce the cognitive load it takes to process a graph!

Show only what you need to show.

Use `theme()` and `element_blank()` to remove different parts of the graph.

---
# The power of `theme()`

Lots of prebuilt themes are really useful!

- `theme_minimal()`

https://ggplot2.tidyverse.org/reference/theme.html

---
# Heatmap Example

```{r}
heat_map <- cvd_race_count %>% 
  ggplot(aes(x=cvd, y=race, fill=count)) + 
  geom_tile(color="black") +
  scale_fill_gradient(low="white", high="lightblue") + 
  geom_text(aes(label=count))
```

---
# Resulting Plot

```{r echo=FALSE}
heat_map
```

---
# Decluttering a heatmap using `theme()`

If you want to remove an element, use `element_blank()`. Here we're also removing the legend with `legend.position = none`, and using `theme_minimal()` to remove some extraneous elements, and resizing the text.

```{r}
heat_map <- cvd_race_count %>% 
  ggplot(aes(x=cvd, y=race, fill=count)) + 
  geom_tile(color="black") +
  scale_fill_gradient(low="white", high="lightblue") + 
  geom_text(aes(label=count), size=10) + #<< 
  theme_minimal() + 
  #remove clutter
  theme(legend.position = "none", #<<
        text = element_text(size=20), #<<
        #remove the x and y axis titles by making them blank
        axis.title.x = element_blank(), #<<
        axis.title.y = element_blank(), #<<
        )
```

---
# Resulting Graph

```{r echo=FALSE}
heat_map
```

---
# Context: Make your categories more understandable

Use of `case_when()` to recode categorical data:

```{r}
cvd_patient %>% 
  mutate(smoking = case_when(smoking == "Y" ~ "smoker", #<<
                             smoking == "N" ~ "non-smoker")) %>% #<<
  mutate(cvd= case_when(cvd == "N" ~ "CVD",
                        cvd == "Y" ~ "None")) %>%
  select(smoking, cvd) %>%
  table()
```

---
# Context: Annotating a graph using `annotate`:

`annotate()` can be extremely powerful in highlighting portions of a plot.

We'll look at an example next.

---
# Full Example: Using Color to Highlight LACE Scores

Showing the distribution of LACE Scores by Admit Source for our cases.

Think of a title for this plot!

```{r}
lace_table <- read.delim("data/analytic-table-LACE.txt")

out_plot <- lace_table %>% 
  mutate(lace7 = ifelse(LACE > 7, "Y", "N")) %>% #<<
  ggplot(aes(y=as.factor(Admit_source), x=LACE, color=lace7)) +
  geom_count() + 
  scale_color_manual(values=c("Y"="orange", "N"="grey")) #<<
```
---
# Resulting Plot

```{r echo=FALSE}
out_plot
```

---
# Making things a little less cluttered

```{r}
lace_table <- read.delim("data/analytic-table-LACE.txt")

out_plot <- lace_table %>% 
  mutate(lace7 = ifelse(LACE > 7, "Y", "N")) %>% 
  ggplot(aes(y=as.factor(Admit_source), x=LACE, color=lace7)) + 
  geom_count() + scale_color_manual(values=c("Y"="orange", "N"="grey")) +
  theme_minimal() + #<<
  geom_vline(xintercept=8, lty=3) + #<<
  theme(legend.position = "none", axis.title.y = element_blank()) #<<
```

---
# Resulting Plot

```{r echo=FALSE}
out_plot
```

---
# Adding Annotation

```{r}
lace_table <- read.delim("data/analytic-table-LACE.txt")

out_plot <- lace_table %>% 
  mutate(lace7 = ifelse(LACE > 7, "Y", "N")) %>% 
  ggplot(aes(y=as.factor(Admit_source), x=LACE, color=lace7)) + 
  geom_count() + scale_color_manual(values=c("Y"="orange", "N"="grey")) +
  theme_minimal() + 
  geom_vline(xintercept=8, lty=3) +
  theme(legend.position = "none", axis.title.y = element_blank()) +  
  annotate("text",x = 5, y= 2.3, label="Low Risk") + #<<
  annotate("text", x=10, y=2.3, label = "High Risk" ) + #<<
  scale_size_area(max_size = 10) #<<
```

---
# The Resulting Plot

```{r}
out_plot
```

---

# Another example: Proportional Barplot 

---
# Context: Subsetting using filter

Think about focusing on your subgroup, or groups to compare.

```{r}
out_plot <- cvd_race_count %>% 
  #using filter to subset the data
  filter(race=="Black/AfAm" | race == "White") %>% #<<
  ggplot(aes(x=race, y=count, fill=cvd)) + 
  geom_bar(stat="identity", position = "fill", color="black") + 
  theme_minimal() + 
  #map Y and N to colors using scale_fill_manual
  scale_fill_manual(values=c("N"= "grey", "Y"="blue")) +
  #remove legend
  theme(legend.position="none") +
  #add an appropriate title
  ggtitle("White males have a higher prevalence of CVD than Black Males")
```

---
# The Plot

```{r echo=FALSE}
out_plot
```

---
# `ggplot2` flipbook

Good examples for styling your plots!

https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html

- [Arctic Ice](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html#226)
- [Flipping Seats](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html#302)
- [Milk Cows](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html#354)

---
# For Much More Info

- Data Visualization Course

---
# Recap

- You are not your audience
- Have an ask, and connect your results to it
- Pick good visuals
- Pare your visuals down and highlight

---
# References

- Source for these slides: https://github.com/laderast/data_storytelling_talk
- [Allison Sliter: Data Presentation for Everyone](https://ww2.amstat.org/meetings/sdss/2019/onlineprogram/AbstractDetails.cfm?AbstractID=306642)
- [ggplot2 flipbook](https://evamaerey.github.io/ggplot_flipbook/ggplot_flipbook_xaringan.html)
- [Storytelling with Data](http://www.storytellingwithdata.com/books)
- Slides are done with xaringan/xaringanthemer
